---
title: "Class 17"
author: "Harshita Jha (PID: A17350910)"
format: pdf
---

## Downstream Analysis

We need to install **tximport** package for straightforward import of Kallisto results. 

- To troubleshoot import, presence of the quant folders were verified using the `dir()` command and the "rhdf5" bioconductor package was installed to read the `abundance.h5` files.

```{r}

dir()

```

```{r}
# BiocManager::install("tximport")
#BiocManager::install("rhdf5")
library(tximport)
library(rhdf5)

folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path( folders, "abundance.h5" )
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```

```{r}
head(txi.kallisto$counts)
```

How many transcripts do we have for each sample? 

```{r}
colSums(txi.kallisto$counts)
```

How many transcripts are detected in at least one sample?

```{r}
sum(rowSums(txi.kallisto$counts)>0)
```

We also have to filter out the annotated scripts with no reads and those with no change over the samples:

```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```


## Principal Component Analysis

```{r}
pca <- prcomp(t(x), scale=TRUE)
summary(pca)
```

### Visualizing the summarized transcriptomic profiles of each sample

```{r}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
```

> Q. Use ggplot to make a similar figure of PC1 vs PC2 and a separate figure PC1 vs PC3 and PC2 vs PC3.

- PC1 vs PC2

```{r}
library(ggplot2)
library(ggrepel)

mycols <- c("blue","blue","red","red")

ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point( col=mycols ) +
  geom_text_repel( col=mycols ) +
  theme_bw()

```


-PC1 vs PC3

```{r}
library(ggplot2)
library(ggrepel)

mycols <- c("blue","blue","red","red")

ggplot(pca$x) +
  aes(PC1, PC3, label=rownames(pca$x)) +
  geom_point( col=mycols ) +
  geom_text_repel( col=mycols ) +
  theme_bw()


```

- PC2 vs PC3 

```{r}
library(ggplot2)
library(ggrepel)

mycols <- c("blue","blue","red","red")

ggplot(pca$x) +
  aes(PC2, PC3, label=rownames(pca$x)) +
  geom_point( col=mycols ) +
  geom_text_repel( col=mycols ) +
  theme_bw()


```

