---
title: "Class 12 Extra Credit"
author: "Harshita Jha (PID: A17350910)"
format: pdf
---

## Population Scale Analysis

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. 

So, you processed about ~230 samples and did the normalization on a genome level. Now, you want to find whether there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.


> Q. How many samples do we have? 

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```


> Q13. Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

```{r}
table(expr$geno)
```

```{r}
box_plot <- boxplot(exp ~geno, data=expr)
```

```{r}
box_plot$stats[3,]
```

- The sample size for A/A is 108, A/G is 233 and G/G is 121. The medians are 31.24847, 25.06486 and 20.07363 respectively. 



> Q14. Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
library(ggplot2)
```

```{r}
ggplot(expr) + aes(x=geno, y=exp, fill=geno) + geom_boxplot(notch=T)
```

- From this plot, we can see that as the number of G alleles increases, the expression decreases so it can be inferred that ORMDL3 expression is highest in A/A, medium in A/G and lowest in G/G.

- This pattern wherein G allele increase is associated with decrease in expression suggests that the SNP does affect ORMDL3 expression in an allele-dependent way. 