---
title: "Class 5: Data Viz with ggplot"
author: "Harshita Jha (PID:A17350910)"
format: pdf
toc: true
---

## Background 

There are lots of ways to make figures in R. These include so-called "base R" graphics (e.g. `plot()`) and tones of add-on packages like **ggplot2**.

For example here we make the same plot with both: 

```{r}
head(cars)
```

```{r}
plot(cars)
```
First I need to install the package with the command `install.packages()`

> **N.B** We never run an install cmd in quarto code chunk or we will end up re-installing packages many many times- which is not we want!

Everytime we want to use one of these "add-on" packages we need to load it up in R with the `library()` function: 

```{r}
library(ggplot2)
```

```{r}
ggplot(cars)
```

Every ggplot needs at least 3 things: 

- The **data**, the stuff we want plotted
- The **aes**thetics, how the data map to the plot
- The **geom**etry, the type of plot

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) + 
  geom_point() 
```

Add a line to better show relationship between the points
```{r}
p <- ggplot(cars) + 
  aes(x=speed, y=dist) + 
  geom_point() + 
  geom_smooth(method = "lm", se=FALSE, color= "green") +
  labs(title="Stopping distance of old cars", subtitle = "Data from the `cars` object", x="Speed (mph)", y="Distance (ft)")
```

Render it out
```{r}
p + theme_bw()
```

## Gene Expression Plot

We can read the input data from the ckass website 

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```
```{r}
nrow(genes)
```
```{r}
ncol(genes)
```

A first version plot 

```{r}
ggplot(genes) + 
  aes(Condition1, Condition2) +
  geom_point()
```

```{r}
table ( genes$State )
```

Version 2 let's color by `State` so we can see the "up" and "down" significant genes compared to all the "unchanging" genes

```{r}
ggplot(genes) + 
  aes(Condition1, Condition2, color=State) +
  geom_point()
```

Version 3 plot, let's modify the default colors to something we like

```{r}
ggplot(genes) + 
  aes(Condition1, Condition2, color=State) +
  geom_point() + 
  scale_color_manual(values=c("red","grey","blue")) + 
  labs(x="Control (no drug)", y= "Drug", title="Gene Expression Changes upon GLP-1") +
  theme_minimal()
```

## Going Further 

Let's have a look at the famous **gapminder** data set

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder, 10)
```
```{r}
ggplot(gapminder) + 
  aes(gdpPercap, lifeExp,col=continent ) +
  geom_point(alpha=0.3)
```

Let's "facet" (i.e make a separate plot) by continent rather than the big hot mess above. 

```{r}
ggplot(gapminder) + 
  aes(gdpPercap, lifeExp,col=continent ) +
  geom_point(alpha=0.3) + 
  facet_wrap(~continent)
```


## Custom Plots

How big is this gapminder dataset? 

```{r}
nrow(gapminder)
```

I want to "filter" down to a subset of this data. I will use the **dplyr** package to help me. 
First I need to install it using `install.packages(dplyr)` and then launch the library using `library(dplyr)`

```{r}
library(dplyr)
```

```{r}
gapminder_2007 <- filter(gapminder, year==2007)
head(gapminder_2007)
```

```{r}
filter(gapminder_2007, country=="Ireland")
```

```{r}
filter(gapminder, year==2007, country=="Ireland")
```

```{r}
filter(gapminder, year== 2007, country== "United States")
```

> Q. Make a plot comparing 1977 and 2007 for all countries

```{r}
input <- filter(gapminder, year %in% c(1977,2007))
```

```{r}
ggplot(input) + aes(gdpPercap, lifeExp, colour = continent) + geom_point() +facet_wrap("year")
```

