---
title: "Class 18 Mini Project"
author: "Harshita Jha (A17350910)"
format: html
---

## Investigating pertussis cases by year



First we have to install the **datapasta** package. 

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
# install.packages("datapasta")
library (datapasta)
library(ggplot2)
cdc <- data.frame(
  year = 1922:2023,
  cases = c(
    107473, 164191, 165418, 152003, 202210, 181411, 161799, 197371,
    166914, 172559, 215343, 179135, 265269, 180518, 147237, 214652,
    227319, 103188, 183866, 222202, 191383, 191890, 109873, 133792,
    109860, 156517, 74715, 69479, 120718, 68687, 45030, 37129,
    60886, 62786, 31732, 28295, 32148, 40005, 14809, 11468,
    17749, 17135, 13005, 6799, 7717, 9718, 4810, 3285,
    4249, 3036, 3287, 1759, 2402, 1738, 1010, 2177,
    2063, 1623, 1730, 1248, 1895, 2463, 2276, 3589,
    4195, 2823, 3450, 4157, 4570, 2719, 4083, 6586,
    4617, 5137, 7796, 6564, 7405, 7298, 7867, 7580,
    9771, 11647, 25827, 25616, 15632, 10454, 13278, 16858,
    27550, 18719, 48277, 28639, 32971, 20762, 17972, 18975,
    15609, 18617, 6124, 2116, 3044, 7063
  )
)
ggplot(cdc) +
  aes(x=year, y=cases) +
  geom_point() +
  geom_line() +
  labs(x="Year", y="Number of cases", title= "Pertussis Cases by Year (1922-2023)" )
```

## A tale of two vaccines (wP & aP)

Examining the CDC data plot to see the changes that happened after the switch to the acellular pertussis (aP) vaccination program 

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc) +
  aes(x=year, y=cases) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 1946, color = "blue", linetype= "dashed") + 
  geom_vline(xintercept = 1996, color = "red", linetype= "dashed") +
  geom_text(aes(x=1946, y=250000, label="wP"),color="blue") + 
  geom_text(aes(x=1996, y=250000, label="aP"), color="red") +
  labs(x="Year", y="Number of cases", title= "Pertussis Cases by Year (1922-2023)" )
```

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

- After the introduction of the aP vaccine, the number of Pertussis cases started to increase over time (as compared to the decreased/low cases before introduction). Even though the latter aP vaccines were developed to have less side effects, since they used only purified antigens of the bacteria instead of the broader 'whole cell' inactivated bacteria, the immune response from the aP vaccine may be more narrow/less lasting as compared to the earlier vaccine, leading to an increase in the number of cases. Other hypotheses include more sensitive PCR-based testing, vaccination hesitancy, bacterial evolution (escape from vaccine immunity) and waning of the immunity in adolescents originally primed as infants with newe aP vaccine as compared to the older wP vaccine.

## Exploring CMI-PB data

### The CMI-PB API returns JSON data

For reading such files, we need to use the `read_json()` from the **jsonlite** package. 

```{r}
library(jsonlite)
```

Read the main subject database table from the CMI-PB API.

```{r}
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
```

```{r}
head(subject, 3)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

- There are 87 aP and 85 wP infancy vaccinated subjects in the dataset.

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

- There are 112 female and 60 male subjects in the dataset.

>Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$biological_sex, subject$race)
```

### Working with dates

We use the **lubridate** package.

```{r}
# install.packages("lubridate")
library(lubridate)
```

Today's date
```{r}
today()
```

How many days have passed since new year 2000?
```{r}
today() - ymd("2000-01-01")
```

What is this in years?
```{r}
time_length( today() - ymd("2000-01-01"),  "years")
```

>Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
library(lubridate)

subject$age <- today() - ymd(subject$year_of_birth)

library(dplyr)

ap <- subject %>% filter(infancy_vac == "aP")

round( summary( time_length( ap$age, "years" ) ) )

```

```{r}
wp <- subject %>% filter(infancy_vac == "wP")
round( summary( time_length( wp$age, "years" ) ) )
```

>Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```

>Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
library(ggplot2)
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

p-value
```{r}
x <- t.test(time_length( wp$age, "years" ),
       time_length( ap$age, "years" ))

x$p.value


```
- From the plot and the p-value, it can be concluded that the two groups are statistically significantly different.

### Joining multiple tables

```{r}
# Complete the API URLs...
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE) 
titer <- read_json("https://www.cmi-pb.org/api/plasma_ab_titer", simplifyVector = TRUE) 
```

>Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
meta <- left_join(specimen, subject)
dim(meta)
head(meta)
```

>Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(titer, meta)
dim(abdata)
```

>Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

>Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

- The most recent dataset (2023 dataset) has the least number of rows (5670), lower than the past few years- it has the least data.


## Examine IgG Ab titer levels

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```


>Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

>Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

- PT, PRN, FIM2/3 and FHA show differences in the level of IgG antibody titers recognizing them over time. These are pertussis-related antigents which are targets of the vaccine/booster response therefore they are expected to be strongly boosted after vaccination. Others are not pertussis-related antigents and therefore are not expected to be strongly boosted after vaccination.


- setting color and/or facet values of the plot to include `infancy_vac` status: 
```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  xlim(0,75) +
  theme_bw()
```

```{r}
igg %>% filter(visit != 8) %>%
ggplot() +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  xlim(0,75) +
  facet_wrap(vars(infancy_vac, visit), nrow=2)
```

> Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
filter(igg, antigen=="OVA") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()


```

```{r}
filter(igg, antigen=="FIM2/3") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

> Q16. What do you notice about these two antigens time courses and the PT data in particular?

- PT levels increase over time and clearly exceed the OVA levels. Their peak is seen around visit 5 and then they decline, this is similar for both the wP and aP subjects.

>Q17. Do you see any clear difference in aP vs. wP responses?

- No, there are no major clear difference in aP vs. wP responses. Both groups show similar overall patterns with PT levels increasing after the boost and then declining afterwards.


- Looking athe the 2021 dataset IgG PT antigen levels over-time:

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

> Q18. Does this trend look similar for the 2020 dataset?

```{r}
abdata.21 <- abdata %>% filter(dataset == "2020_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2020 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

- Yes, the trend looks similar overall as the IgG PT levels increase after the boost and then decline afterwards but it should be noted that between the 2021 and 2020 datasets- 2020 data is noisier and appears more scattered so the trend is less clean as compared to the cleaner 2021 data set.

## Obtaining CMI-PB RNASeq data

```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"

rna <- read_json(url, simplifyVector = TRUE) 
```

```{r}
ssrna <- inner_join(rna, meta)
```

>Q19. Make a plot of the time course of gene expression for IGHG1 gene (i.e. a plot of visit vs. tpm).

```{r}
ggplot(ssrna) +
  aes(visit, tpm, group=subject_id) +
  geom_point() +
  geom_line(alpha=0.2)
```

> Q20.: What do you notice about the expression of this gene (i.e. when is it at it’s maximum level)?

- The expression of this gene is at its maximum level arount visit 4 (lower at earlier visits) and then it decreases again after visit 4.

> Q21. Does this pattern in time match the trend of antibody titer data? If not, why not?

- This pattern does not exactly match the antibody titer data which showed longer elevation while this patterns shows that it peaks around visit 4 and then declines afterwards. This may be possibly due to the fact that antibody levels can stay remain even after gene expression has declined therefore we see elevated titer data levels while this gene expression shows earlier decline.


```{r}
ggplot(ssrna) +
  aes(tpm, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(vars(visit))
```

```{r}
ssrna %>%  
  filter(visit==4) %>% 
  ggplot() +
    aes(tpm, col=infancy_vac) + geom_density() + 
    geom_rug() 
```

